Mammalian protein-coding genes exhibit widespread beneficial mutations that are not adaptive

### 2023-03-13 manuscript/main.tex
Version for corrections.

\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

@@

\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{xfrac, nicefrac}
\usepackage{bm}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{graphicx}
\usepackage{bbold}
\usepackage{pdfpages}
\pdfinclusioncopyfonts=1
\usepackage{lineno}

@@

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\textrm{Var}}
\newcommand{\der}{\textrm{d}}
\newcommand{\e}{\textrm{e}}
\newcommand{\avg}[1]{\left< #1 \right>} % for average
\newcommand{\Ne}{N_{\textrm{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\textrm{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S_{0}}
\newcommand{\Sphyclass}{\mathcal{C}}
\newcommand{\SphyMean}{\overline{\Sphy}}
\newcommand{\divStrongDel}{\Sphy < -3}
\newcommand{\divDel}{-3 < \Sphy < -1}
\newcommand{\divWeakDel}{-1 < \Sphy < 0}
\newcommand{\divWeakAdv}{0 < \Sphy < 1}
\newcommand{\divAdv}{ \Sphy > 1}
\newcommand{\PdivStrongDel}{\proba \left[ \divStrongDel \right]}
\newcommand{\PdivDel}{\proba \left[ \divDel \right]}
\newcommand{\PdivWeakDel}{\proba \left[ \divWeakDel \right]}
\newcommand{\PdivWeakAdv}{\proba \left[ \divWeakAdv \right]}
\newcommand{\PdivAdv}{\proba \left[ \divAdv \right]}
\newcommand{\given}{\mid}
\newcommand{\Spop}{S}
\newcommand{\SpopMean}{\overline{\Spop}}
\newcommand{\polyDel}{\Spop < -1}
\newcommand{\polyNeutral}{-1 < \Spop < 1}
\newcommand{\polyAdv}{ \Spop > 1}
\newcommand{\PpolyDel}{\proba \left[ \polyDel \right]}
\newcommand{\PpolyNeutral}{\proba \left[ \polyNeutral \right]}
\newcommand{\PpolyAdv}{\proba \left[ \polyAdv \right]}
\newcommand{\AdvMean}{\beta_b}
\newcommand{\DelMean}{\beta_d}
\renewcommand{\baselinestretch}{1.5}
\linenumbers

\title{Up to 25\% of beneficial mutations in human protein sequences are not adaptive innovations in mammals}

\author{
    \large
    \textbf{T. {Latrille}$^{1}$, J. {Joseph}$^{2}$, D.~A. {Hartasánchez}$^{1}$, N. {Salamin}$^{1}$}\\
    \normalsize
    $^{1}$Université de Lausanne, Lausanne, Switzerland\\
    $^{2}$Université de Lyon, CNRS, LBBE UMR 5558, Villeurbanne, France \\
    \texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}

@@

%TC:endignore
    \begin{abstract}
        Mutations can be beneficial by bringing an innovation to their bearer, allowing them to adapt to changes in environments or in selective pressure. However, mutations can also be beneficial because they are repairing previous deleterious changes, simply restoring existing functions. In this study, we first estimated selective effects of mutations inside mammalian protein coding sequences, under a model assuming no adaptation at the phylogenetic scale. We then estimated the proportion of beneficial mutations that are not adaptive innovations among all beneficial mutations at the population scale. Our work confirms that deleterious substitutions have accumulated in mammals and are currently being eliminated. In modern humans, it results in around 25\% of beneficial mutations that are not adaptive innovations, but instead are repairing previous deleterious changes.
    \end{abstract}

    \keywords{Adaptation \and Back-mutation \and Phylogenetics \and Population-genetics \and Codon models }

    Adaptation is considered one of the main processes shaping the diversity of forms and functions across the tree of life~\cite{darwin_origin_1859}.
    It enables species to access new ecological niches and also respond to changes in their environment~\cite{darwin_origin_1859} [NS: maybe cite Losos or some key papers on ecological speciation rather than Darwin twice].
    In evolution, adaptation is tightly linked to the notion of change at the level of the environment and to species responding to this change~\cite{merrell_adaptive_1994}.
    The process of adaptation leaves marks of accelerated evolution in genomes since mutations that are beneficial under new conditions fix faster than neutral mutations (fig.~\ref{fig:fitness-landscape}A).
    The signature of an accelerated rate of substitutions, defined as the rate of mutation times the probability of fixation, can thus be interpreted as a sign of adaptation~\cite{mcdonald_adaptative_1991, smith_adaptive_2002, welch_estimating_2006}.
    The current availability of large-scale genomic data and the development of theoretical models have allowed to detect and quantify adaptation across genes and lineages, and these analyses have become common practice in evolutionary biology~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019} [NS: cite the paper by Anna Marcionetti here].
    Although these approaches have led to a better understanding of the processes determining the rates of molecular evolution, they have had the effect of associating beneficial mutations with adaptive evolution, while disconnecting them from the notion of change defining adaptation.
    We argue that we should limit the use of adaptive mutations to those that are associated with adaptation as such and dissociate beneficial mutations from adaptive mutations since adaptive evolution is not the only process that can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009}.
    We propose here a new approach to identify the mechanisms underlying the origin of beneficial mutations.
    In particular, we suggest that many beneficial mutations are not adaptive, but are rather restoring ancestral states of higher fitness.
    We used large-scale genomic data to integrate analyses at both the phylogenetic and population scales to differentiate between truly adaptive mutations from mutations restoring the ancestral fitness.
    Our results show that up to a quarter of beneficial mutations across mammal species are not driven by adaptive innovations, but rather compensate the effect of ancestral mutations to recover the optimal state of the species.

    In a constant environment, if a population is already adapted to these conditions, a deleterious mutation in an individual can reach fixation in the population by genetic drift~\cite{Ohta1992}.
    Subsequently, a new mutation that is restoring the ancestral fitness will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment occurred~\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
    The restoration of the ancestral fitness can happen at a different locus than the initial mutation, in which case it is referred to as a compensating [NS: ou compensatory ?] mutation~\cite{hartl_compensatory_1996, mustonen_fitness_2009}.
    However, if the restoration of the ancestral fitness happens at the same locus than the initial mutation, in which case it is referred to as a beneficial back-mutation~\cite{piganeau_estimating_2003, charlesworth_other_2007}.
    In this study, we focus on the latter case.
    While compensatory mutations induce genetic changes at the sequence level, and thus genetic diversification, beneficial back-mutations reduce the genetic diversity and do not contribute to genetic innovation.
    They merely cancel the effect of the previous deleterious changes.
    As a consequence, not all beneficial mutations contribute to genetic innovation and truly adaptive evolution can only be correctly estimated if we can account for the number of beneficial back-mutations~\cite{keightley_what_2010, rice_evolutionarily_2015}.
    % Philosophically, our distinction between beneficial back-mutations and adaptation can appear to be far-fetched, and one can argue that beneficial back-mutations is adaptation to its own deteriorating genome.
    % However, there is a difference between adapting to a mutation, for instance by means of a compensatory mutation elsewhere, and cancelling the mutation, by means of a substitution on the misfit amino-acid.

    %For an individual carrying a beneficial mutation, whether this mutation is a new innovation responding to a change in environment or a beneficial back-mutation is indistinguishable~\cite{charlesworth_other_2007}.
    Given a beneficial mutation, distinguishing whether it is a new innovation responding to a change in environment or a beneficial back-mutation is impossible with only one individual at hand~\cite{charlesworth_other_2007}.
    This limitation persists at the level of the population, since both types of mutation will result in a positive transmission bias of the beneficial allele.
    However, at the macro-evolutionary scale the consequences of these two types of mutations are fundamentally different.
    While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promote phenotype stability and preserve well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    Additionally, the direction of adaptive evolution is unpredictable because it is caused by an unpredictable change in the environment and, hence, in the underlying fitness landscape~\cite{bazykin_changing_2015}.
    On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known because we will expect to see changes from non-optimal amino acids to optimal ones.
    If we wish to quantify the amount of beneficial back-mutations that have happened along a phylogeny we need to identify the selection coefficients of mutations and to achieve this we require the fitness landscape to be stable and be known.

    The placental mammals represent an excellent study system to perform this analysis.
    Having originated ~100 Mya\cite{kumar_timetree_2017}, they diversified quickly.
    Additionally, we have polymorphism data available for many species\cite{howe_ensembl_2021} and high quality protein-coding DNA alignments across the genome\cite{ranwez_orthomam_2007, scornavacca_orthomam_2019}.
    By restricting our analysis to protein-coding orthologous genes to all the phylogeny, and further excluding any genes for which there have been claims of pervasive directional selection, we can not only assume a constant fitness landscape for each of them throughout placental mammal evolution, but actually reconstruct the fitness landscape from variable positions within the sequence along the phylogeny.
    For each gene, we used a codon model of mutation-selection~\cite{halpern_evolutionary_1998, mccandlish_modeling_2014} to obtain relative fitnesses for all amino-acids for each site of the sequence by fitting the model to the multi-species sequence alignment (fig~\ref{fig:method}A) and assuming that the underlying fitness landscape is stable along the phylogenetic tree.
    %Mutation-selection codon models provide a nearly neutral framework of evolution taking into account non-adaptive processes impacting codon frequency~\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022}.
    From the fitness landscape, we can calculate the scaled selection coefficient $\Sphy$ for any possible amino-acid change at each position.
    Under this model, a mutation with a positive $\Sphy$ would be driving the current site toward a known fitter amino-acid and would thus be considered as a beneficial back-mutation (fig~\ref{fig:method}B), given the assumed stability of the fitness landscape.

    Having identified which potential amino-acid changes could represent beneficial back-mutations at the phylogenetic scale, we retrieved polymorphism data from 28 populations belonging to 6 genera (\textit{Equus}, \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus} and \textit{Homo}) to assess the presence of beneficial back-mutations at the population scale.
    We focused on currently segregating mutations within populations and also on substitutions in the terminal lineages, and checked if any of these observed changes where predicted beneficial back-mutations.
    Essentially, by integrating large scale genomic datasets at both phylogenetic and population scales we have been able to estimate the amount of beneficial back-mutations across the entire exome for 6 different mammalian genera (fig~\ref{fig:method}A-F).

    Our analyses allow us to test long-standing questions in molecular evolution:
    How common are beneficial back-mutations?
    Can we identify beneficial back-mutations that have reached fixation recently, and ones that are currently segregating in populations?
    Are hypothetical beneficial back-mutations effectively beneficial for the individuals carrying them?
    Furthermore, what is the probability for any beneficial mutation observed in current populations to be a back-mutation?
    In other words, can we quantify the amount of beneficial mutations that are restoring damaged genomes instead of creating adaptive innovations.
    By addressing these questions we provide
    novel insight on how selection is affecting protein-coding DNA sequences.

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.fitness-landscape}
        \caption{
            Panel A and B: at a given site of a protein-coding DNA sequence, the different amino acids (x-axis) have different fitnesses (y-axis).
            Under a fluctuating fitness landscape (panel A), these fitnesses are changing with time.
            The protein sequence is always lagging behind the moving target defined by the amino acid fitnesses, and since substitutions are accepted preferentially if they are in the direction of this target, substitutions are on average adaptive.
            At the phylogenetic scale (panel C), this phenomena promotes phenotype diversification across species.
            Under a stable fitness landscape (panel B), all the mutations reaching fixation are either slightly deleterious and reaching fixation due to drift, or are beneficial back-mutations restoring a more optimal amino acid.
            At the phylogenetic scale (panel D), this phenomena promotes promotes phenotype stability and preserves well established biological systems.
            At the genome level, slightly deleterious mutations that have reached fixation are supposedly scattered at different loci, such that beneficial back-mutations should also be spread across the genome.
            We thus expect that the genome-wide signature of beneficial back-mutations can be detected and quantified, even though each mutation has individually a small beneficial effect on their bearer.
        }
        \label{fig:fitness-landscape}
    \end{figure*}

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.method.proba}
        \caption{
            Integrating datasets at different scales.
            At the phylogenetic level (panel A), the amino acid Wrightian fitness for each site is estimated from protein-coding DNA alignments using mutation-selection codon models.
            Mutation-selection models assume a stable fitness landscape, taking into account only non-adaptive evolution.
            A mutation with a $\Sphy > 1$ would be driving the current site toward are fitter amino-acid, and such mutation are considered a beneficial back-mutation repairing existing functions (panel B).
            At the population scale (panel C), each single nucleotide polymorphism (SNP) segregating in the population can be classified as a beneficial back-mutation (in blue) or not.
            The number of derived alleles allows us to estimate the probability of beneficial mutations at the population scale, for all SNPs or among back-mutations (panel D).
            The probability of back-mutation is obtained as their fraction among all possible mutations from the ancestral genome sequence (panel E).
            Altogether, we can estimate the probability of back-mutations among all beneficial ones, using Bayes' theorem (panel F).
        }
        \label{fig:method}
    \end{figure*}

    \subsection*{Back-mutations in the terminal lineages}
    We first estimated the number of beneficial back-mutation in the terminal branches of mammals by reconstructing the substitutions occurring along the terminal branches of the phylogenetic tree.
    For each substitution, we estimated its scaled selection coefficient ($\Sphy$) based on the stable fitness landscape of amino-acids estimated at the scale of the phylogenetic tree of mammals (fig.~\ref{fig:method}A).
    Among all the substitutions that were found in the terminal branches, between 10 and 13\% had $\Sphy > 1$ (table~S1), while between 25 and 30\% had $\Sphy > 0$ (table~S2).
    For substitutions with $\Sphy > 1$, the $\dnds$ ratio of non-synonymous over synonymous divergence was estimated to be between 1.15 and 1.68 in the different lineages (table~S1).
    Beneficial back-mutations with $\Sphy > 1$ therefore reached fixation more frequently than synonymous mutations that are supposed to be neutral, which means that these back-mutations are effectively beneficial.
    This result further indicated that $\dnds$ used to detect adaptation while relying on substitutions between closely related species~\cite{mcdonald_adaptative_1991, galtier_adaptive_2016} or at the phylogenetic scale~\cite{goldman_codonbased_1994, yang_codonsubstitution_2002} is biased due to the presence of beneficial back-mutations that are not adaptive.
    We can alleviate this bias by computing $\dnds$ exclusively on non-synonymous sites with a $\Sphy < 0$, which represents the rate of evolution when beneficial back-mutations are discarded.
    By comparing these two ways of estimating $\dnds$ (see Material \& Methods), we estimated that between 22\% and 27\% of $\dnds$ is inflated because of beneficial back-mutations while they only represent between 1\% and 2\% of mutational opportunities (table~S2).

    \subsection*{Back-mutations in the populations}
    We then estimated the proportion of new mutations that were beneficial back-mutations while being under selection at the population scale (fig.~\ref{fig:method}C-E).
    We first reconstructed the ancestral genome (instead of the reference genome) of each mammal species with available polymorphism data.
    We computed the scaled selection coefficient ($\Sphy$) along this ancestral genome for each possible non-synonymous mutations, weighted by their mutational opportunity (see Material \& Methods).
    It allowed us to obtain the distribution of fitness effects (DFE) of new mutations from the ancestral sequence under the assumption of the stable fitness landscape as shown for \textit{Homo sapiens} in Fig.~\ref{fig:homo-afr-results}A.
    Because $\Sphy$ was based on estimation of fitnesses at the mammalian scale, mutations with $\Sphy>1$ were considered, as shown above, beneficial back-mutations and only represented ca.~1.5\% of all possible new mutations (blue in fig.~\ref{fig:homo-afr-results}A, table~S3).
    However, when focusing on the mutations that are observed in populations, namely single nucleotide polymorphism (SNP), beneficial back-mutations represents ca.~8.5\% of all observed mutations (blue in fig.~\ref{fig:homo-afr-results}B, table~S3).

>>>>

\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

@@

\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{xfrac, nicefrac}
\usepackage{bm}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{graphicx}
\usepackage{bbold}
\usepackage{pdfpages}
\pdfinclusioncopyfonts=1
\usepackage{lineno}
\usepackage{tabu}
\usepackage{hhline}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\usepackage{booktabs}
\usepackage{makecell}

@@

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\der}{\text{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S_{0}}
\newcommand{\Sphyclass}{\mathcal{C}}
\newcommand{\given}{\mid}
\newcommand{\Spop}{S}
\newcommand{\PpolyDel}{\proba [ \Spop^{\text{d}}]}
\newcommand{\PpolyNeutral}{\proba [ \Spop^{\text{n}} ]}
\newcommand{\PpolyAdv}{\proba [ \Spop^{\text{b}} ]}
\newcommand{\AdvMean}{\beta_b}
\newcommand{\DelMean}{\beta_d}
\renewcommand{\baselinestretch}{1.5}
\renewcommand{\arraystretch}{0.6}
\linenumbers
\frenchspacing

\title{Mammalian protein-coding genes exhibit widespread beneficial mutations that are not adaptive}
% Empirical evidence for widespread beneficial mutations that are not adaptive in mammalian genes

\author{
    \large
    \textbf{T. {Latrille}$^{1}$, J. {Joseph}$^{2}$, D.~A. {Hartasánchez}$^{1}$, N. {Salamin}$^{1}$}\\
    \normalsize
    $^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
    $^{2}$Université de Lyon, Université Lyon 1, CNRS, VetAgro Sup, Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Villeurbanne, France \\
    \texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}

@@

%TC:endignore
    \begin{abstract}
        Mutations can be beneficial by bringing an innovation to their bearer, allowing them to adapt to changes in their environments. However, mutations can also be beneficial because they are reverting previous deleterious changes, simply restoring pre-existing functions. By integrating phylogenomic and population data in mammals we estimated the contribution of beneficial back-mutations to molecular evolution, which has remained widely overlooked to this day. Applying a mutation-selection model to mammalian conserved protein-coding genes, we estimated amino-acid fitness landscapes. Assuming no changes in these fitness landscapes, any beneficial mutation is necessarily a back-mutation. We confirm that these beneficial back-mutations are positively selected in extant populations and demonstrate that a substantial part of ongoing positive selection is not driven by adaptation to environmental change in mammals.
    \end{abstract}

    \keywords{Adaptation \and Back-mutations \and Phylogenetics \and Population-genetics \and Codon models }

    Adaptation is considered one of the main processes shaping the diversity of forms and functions across the tree of life\cite{darwin_origin_1859}.
    In evolution, adaptation is tightly linked to the notion of environmental change and to species responding to this change\cite{merrell_adaptive_1994, gavrilets_adaptive_2009}.
    For adaptation to take place, there needs to be variation within populations, which at the molecular level mostly appears via mutations in the DNA sequence.
    While neutral mutations will have no impact on the fitness of an individual, deleterious mutations have a negative effect and beneficial mutations lead their bearer toward a higher fitness.
    Compared to a neutral mutation, a beneficial mutation is thus more likely to reach fixation in the population, resulting in a substitution at the species level.
    Upon environmental change, adaptive beneficial mutations toward new fitness optima are more likely to rise in frequency, resulting in an increased substitution rate (fig.~\ref{fig:fitness-landscape}A).
    Altogether, an accelerated substitution rate is thus commonly interpreted as a sign of adaptation\cite{mcdonald_adaptative_1991, smith_adaptive_2002, welch_estimating_2006}.
    The current availability of large-scale genomic data and the development of theoretical models have enabled us to detect and quantify acceleration of substitution rates across genes and lineages, and these analyses have become common practice in evolutionary biology\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
    These approaches have led to a better understanding of the processes determining the rates of molecular evolution.
    However, they have had the collateral effect of associating beneficial mutations with adaptive evolution when, in fact, adaptive evolution is not the only process that can lead to beneficial mutations\cite{charlesworth_other_2007, mustonen_fitness_2009}.

    In a constant environment, a deleterious mutation can reach fixation in the population by genetic drift\cite{Ohta1992}.
    Subsequently, a new mutation restoring the ancestral fitness will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment has occurred\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
    The restoration of the ancestral fitness can happen with a mutation occurring either at a different locus from that of the initial mutation, in which case it is referred to as a compensatory mutation\cite{hartl_compensatory_1996, mustonen_fitness_2009}, or at the same locus of the initial mutation, referred to as a beneficial back-mutation\cite{piganeau_estimating_2003, charlesworth_other_2007}.
    Now, while compensatory mutations induce genetic changes at the sequence level, and thus genetic diversification, beneficial back-mutations reduce the inter-specific genetic diversity and do not contribute to genetic innovation.
    Although beneficial back-mutations were considered negligible in the enunciation of the nearly-neutral theory by Tomoko Ohta\cite{Ohta1992}, their importance during a population size expansion has been acknowledged more recently\cite{charlesworth_other_2007}.
    However, differentiating between an adaptive mutation and a beneficial back-mutation has remained challenging\cite{chi_detecting_2020}.
    In this study, we seek to find empirical evidence for beneficial back-mutations and to test long-standing questions in molecular evolution such as how common are beneficial back-mutations, and
    what is the probability for any beneficial mutation observed in current populations to be a back-mutation.
    By addressing these questions we provide novel insight on the process of adaptation and, more broadly, on how natural selection has affected protein-coding DNA sequences throughout evolution.

    \subsection*{Fitness landscape reconstruction}

    An adaptive mutation responding to a change in the environment or a beneficial back-mutation have equivalent consequences for their bearer\cite{charlesworth_other_2007}.
    Similarly, at the level of the population, both types of mutations will result in a positive transmission bias of the beneficial allele.
    However, at the macro-evolutionary scale the consequences of these two types of mutations are fundamentally different.
    While adaptive mutations promote phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promote phenotype stability and may help preserve well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    Additionally, the direction of adaptive evolution is unpredictable because it is caused by an unpredictable change in the environment and, hence, in the underlying fitness landscape\cite{bazykin_changing_2015}.
    On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known because changes from non-optimal amino acids to optimal ones will be expected.
    In this context, the mutation-selection framework permits to relate the patterns of substitution along a phylogeny to the underlying fitness landscape\cite{halpern_evolutionary_1998, mccandlish_modeling_2014}.
    Such mutation-selection models applied to protein coding DNA sequences allow us to estimate relative fitnesses for all amino-acids for each site of the sequence, explicitly assuming that the underlying fitness landscape is stable along the phylogenetic tree\cite{rodrigue_mechanistic_2010, tamuri_estimating_2012, rodrigue_detecting_2017}.
    Importantly, because mutation-selection models at the phylogenetic scale are based on population-genetics equations, their estimates of fitness effects are directly interpretable as fitness effects for an individual within a population.
    %These methods account for phylogenetic relationships, the structure of the genetic code and accommodate for underlying mutational biases\cite{rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022}.

    Theoretically, we can thus use large-scale genomic data to assess whether the fitnesses estimated at the phylogenetic scale are predictive of the fitness effect at the population scale, for both deleterious and beneficial back-mutations.
    The placental mammals represent an excellent study system to perform such analysis.
    Having originated $\sim$100 Mya\cite{kumar_timetree_2017}, they diversified quickly.
    Additionally, polymorphism data are available for many species\cite{howe_ensembl_2021} as high quality protein-coding DNA alignments across the genome\cite{ranwez_orthomam_2007, scornavacca_orthomam_2019}.
    By restricting our analysis to protein-coding orthologous genes across all the phylogeny, we restrict our analysis to genes mostly under functional constrains across mammals.
    For each gene, fitting the mutation-selection model to the multi-species sequence alignment assuming that the underlying fitness landscape is stable along the phylogenetic tree allows obtain relative fitnesses for all amino-acids for each site of the alignment (fig~\ref{fig:method}A).
    From the fitness landscape, we can calculate the scaled selection coefficient $\Sphy$ for any possible amino-acid change at each position.
    From the selection coefficient, we can classify any mutation as either a deleterious mutation toward a less fit amino-acid ($\Sphy^{\text{d}}$), a nearly-neutral mutation ($\Sphy^{\text{n}}$) or a beneficial back-mutation toward a known fitter amino-acid ($\Sphy^{\text{b}}$).
    Having identified which potential DNA changes represent beneficial back-mutations (fig~\ref{fig:method}B), we retrieved polymorphism data from 28 populations belonging to 6 genera (\textit{Equus}, \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus} and \textit{Homo}) to assess the presence of beneficial back-mutations at the population scale.
    We focused on currently segregating mutations within populations and also on substitutions in the terminal branches, and checked if any of these observed changes where beneficial back-mutations (fig~\ref{fig:method}C-E).
    Essentially, by integrating large scale genomic datasets at both phylogenetic and population scales we have been able to estimate the amount of beneficial back-mutations across the entire exome for 6 different mammalian genera (fig~\ref{fig:method}F-H).

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.fitness-landscape}
        \caption{
            (A \& B) At a given site of a protein-coding DNA sequence, the different amino-acids (x-axis) have different fitness values (y-axis).
            Under a fluctuating fitness landscape (A), these fitnesses change with time.
            The protein sequence follows the moving target defined by the amino-acid fitnesses, and since substitutions are accepted preferentially if they are in the direction of this target, substitutions are on average adaptive.
            At the phylogenetic scale (C) beneficial substitutions are common (positive signs), promoting phenotype diversification across species.
            Under a stable fitness landscape (B), most mutations reaching fixation are either slightly deleterious reaching fixation due to drift, or are beneficial back-mutations restoring a more optimal amino-acid.
            At the phylogenetic scale (D) deleterious substitutions (negative signs) are often reverted via beneficial back-mutations (positive signs), promoting phenotype stability and preserving well established biological systems.
            Even though individually any back-mutation might have a small beneficial effect on its bearer, we expect beneficial back-mutations to be scattered across the genome and the genome-wide signature of beneficial back-mutations to be detectable and quantifiable.}
        \label{fig:fitness-landscape}
    \end{figure*}

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.method.proba}
        \caption{
            Selection coefficients at the phylogenetic and population scales.
            At the phylogenetic scale (A), the amino-acid Wrightian fitness for each site is estimated from protein-coding DNA alignments using mutation-selection codon models.
            For every possible mutation, the difference in amino-acid fitness before and after the mutation allows us to compute the selection coefficient at the phylogenetic scale ($\Sphy$).
            Mutations with $\Sphy < -1$ are considered deleterious ($\Sphy^{\text{d}}$ in red), and mutations with $-1 < \Sphy < 1$ are considered nearly-neutral ($\Sphy^{\text{n}}$ in green).
            Mutations with $\Sphy > 1$ are toward a fitter amino-acids, and thus are considered beneficial back-mutations ($\Sphy^{\text{b}}$ in green blue), repairing existing functions (B).
            At the population scale (C), each observed single nucleotide polymorphism (SNP) segregating in the population is classified according to its $\Sphy$ value.
            Similarly, substitutions along the terminal branch are classified depending on their $\Sphy$ value.
            Together, substitutions and polymorphisms are contrasted to synonymous changes used as a neutral expectation such as estimate selection coefficients at the population scale ($\Spop$).
            For each class of selection ($\Sphy^{\text{b}}$, $\Sphy^{\text{n}}$, $\Sphy^{\text{d}}$), we can thus assess whether $\Sphy$ predicts $\Spop$ (E), and compute \textit{precision} (F-G) and \textit{recall} (F-H) for each class.
            The \textit{recall} value for the class $\Sphy^{\text{b}}$ is the probability of back-mutations among all beneficial ones (H).
        }
        \label{fig:method}
    \end{figure*}

    \subsection*{Selection along the terminal branches}
    First, we assessed whether fitness effects derived from the mutation-selection model at the phylogenetic scale are predictive of selection occurring in terminal branches.
    To achieve this, we recovered the mutations that reached fixation in the terminal branches of 6 different mammalian genera.
    Only mutations fixed in a population were considered as substitutions in the corresponding branch, while segregating mutations were discarded from this analysis.
    For each substitution identified in the terminal branches, we obtained its scaled selection coefficient ($\Sphy$) as the change in amino-acid fitness at the corresponding codon site in the mammalian alignment (fig.~\ref{fig:method}A-C).
    Importantly, the mammalian alignment used to estimate amino-acid fitness landscape does not include the genera for which we reconstructed the substitutions, to avoid fallacy and circularity in the reasoning.
    Because $\Sphy$ was based on estimation of fitness effects at the mammalian scale, mutations with $\Sphy>1$ were considered beneficial back-mutations ($\Sphy^{\text{b}}$), toward a fitter amino acid.
    Among all the substitutions that were found in each terminal branch, between 10 and 13\% were $\Sphy^{\text{b}}$ (table~S1).
    In principle, beneficial back-mutations are bound to reach fixation more often than neutral substitutions.
    Hence, we calculated the $\dnds$ ratio of non-synonymous over synonymous divergence for all terminal lineages, restricting the non-synonymous changes to our set of $\Sphy^{\text{b}}$.
    We obtained values between 1.15 and 1.68 in the different lineages (table~S1), implying that $\Sphy^{\text{b}}$ reached fixation more frequently than synonymous mutations that are supposed to be neutral, which means that these back-mutations are effectively beneficial.

    This result further indicated that using $\dnds$ as a proxy for the efficiency of purifying selection is biased due to the presence of beneficial back-mutations.
    By discarding beneficial back-mutations from the estimation of $\dnds$, this corrected estimation represents the true efficiency of purifying selection.
    By comparing these two ways of calculating $\dnds$ (see Materials \& Methods), we estimated that $\dnds$ values are inflated by between 22\% and 27\% across genera because of beneficial back-mutations while the latter only represent between 1\% and 2\% of mutational opportunities (table~S2).

    \subsection*{Selection in populations}
    Second, we assessed whether our scaled selection coefficients ($\Sphy$) were predictive of selective forces exerted at the population level.
    To achieve this, we reconstructed the ancestral genome of each population to polarize single nucleotide polymorphism (SNPS) segregating in the population.
    The scaled selection coefficient ($\Sphy$) predicted for every non-synonymous SNPs allow us to classify them as either $\Sphy^{\text{d}}$, $\Sphy^{\text{n}}$ or $\Sphy^{\text{b}}$ (fig.~\ref{fig:method}B-C).

    In humans, some SNPs have been associated to specific clinical prognosis terms\cite{landrum_clinvar_2018}. Those clinical terms have been obtained by clinical evaluation of the impact of variants on human Mendelian disorders\cite{landrum_clinvar_2018}. Thus they provide an orthogonal estimate of the effect of a mutation on human health, giving us an opportunity to evaluate whether this effect is in accordance with its predicted fitness effect.
    To do so, we investigated whether the non-synonymous SNPs classified as $\Sphy^{\text{d}}$ or as $\Sphy^{\text{b}}$ showed an enrichment in specific clinical terms in comparison to SNPs classified as $\Sphy^{\text{n}}$.
    Our results show that SNPs predicted as deleterious are associated to clinical terms such as \textit{Likely Pathogenic} and \textit{Pathogenic}, implying that, in general, the selective pressure of a mutation exerted across mammals is also predictive of its clinical effect in humans (table~S6).
    Conversely, back-mutations are associated with clinical terms such as \textit{Benign} and \textit{Likely Benign}, which shows that back-mutations are less likely to be functionally damaging.
